---
title: "NYPD Shooting Incident Analysis"
author: "Massimo Cuomo"
date: "2024-06-25"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load Libraries
```{r library, echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(sf)
```

Load Dataset
```{r data, echo=TRUE}
data <- read.csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
```

```{r tidy}
# Tidy the data
data <- data %>%
  rename_all(tolower) %>%
  rename(
    date = occur_date,
    time = occur_time,
    borough = boro
  ) %>%
  mutate(
    date = as.Date(date, format="%m/%d/%Y"),
    time = hms::as_hms(time),
    hour = hour(time),
    year = year(date),
    month = month(date),
    day_of_week = weekdays(date)
  )
```

```{r intro}
# View Cleaned Dataset Structure
str(data)
```

```{r geography}
# Remove rows with missing latitude or longitude
data_clean <- data %>%
  filter(!is.na(latitude) & !is.na(longitude))

# Convert to spatial data frame
data_sf <- st_as_sf(data_clean, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)

# Plot the geographic distribution
ggplot(data_sf) +
  geom_sf() +
  labs(title = "Geographic Distribution of Shooting Incidents")
```

```{r timing}
# Plot incidents by time of day
data_hourly <- data %>%
  mutate(hour = hour(time)) %>%
  group_by(hour) %>%
  summarise(incidents = n())

ggplot(data_hourly) +
  geom_line(aes(x = hour, y = incidents)) +
  labs(title = "Shooting Incidents by Time of Day", x = "Time", y = "Incidents")

# Plot incidents by day of the week
ggplot(data) +
  geom_bar(aes(x = day_of_week), fill = "steelblue") +
  labs(title = "Shooting Incidents by Day of the Week", x = "Day", y = "Incidents")

# Plot incidents by year
ggplot(data) +
  geom_bar(aes(x = factor(year)), fill = "steelblue") +
  labs(title = "Shooting Incidents by Year", x = "Year", y = "Incidents")
```

```{r model, message=FALSE}
# Aggregate incident data by perpetrator race and sex
data_model <- data %>%
  filter(perp_race != "", !is.null(perp_race), perp_sex != "", !is.null(perp_sex)) %>%
  group_by(perp_race, perp_sex) %>%
  summarise(incidents = n()) %>%
  ungroup()
# Create model
model <- lm(incidents ~ perp_race + perp_sex, data = data_model)
```

```{r model_visual, fig.height = 7.5, fig.width = 10}
# Predict incidents based on the model
data_model$predicted_incidents <- predict(model, newdata = data_model)

# Plot actual vs predicted incidents
ggplot(data_model) +
  geom_point(aes(x = incidents, y = predicted_incidents, color = perp_race), size = 3) +
  geom_text(aes(x = incidents, y = predicted_incidents, label = perp_sex), vjust = -1, size = 4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(title = "Shooting Incidents by Race and Sex", x = "Actual Incidents", y = "Predicted Incidents") +
  theme_minimal()
```

### Possible Biases
* There is no data in this dataset on the socioeconomic status of perpetrators (income) or of the neighborhoods in which the shooting incidents took place. There could potentially be a large correlation with income.
* There could be possible biases in the police apprehension of suspects of a particular race, or patrolling neighborhoods in which those races are more prevalent.
* There is no data regarding the context of the shootings, whether they were done criminally or potentially in self-defence scenarios.
* There are some incidents with race labeled as "unknown" or "null", potentially skewing the data.
* Perhaps my own bias led me to perform this analysis, based on media coverage and hearsay of social issues in New York.

